---
title: "Hypothesis testing"
subtitle: "<br><br> Data Science in a Box"
author: "[datasciencebox.org](https://datasciencebox.org/)"
output:
  xaringan::moon_reader:
    css: ["../xaringan-themer.css", "../slides.css"]
    lib_dir: libs
    anchor_sections: FALSE
    nature:
      ratio: "16:9"
      highlightLines: true
      highlightStyle: solarized-light
      countIncrementalSlides: false
---

```{r child = "../setup.Rmd"}
```

```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(tidymodels)
library(DT)
#library(emoji)
library(openintro)
```

class: middle

# 

---

## Packages

```{r message=FALSE}
library(tidyverse)
library(tidymodels)
```


---
## Hypothesis test

### Medical example

- **Fact for the purposes of this class:** it is proven that the probability of serious complications arising is, for any hospital in Portugal, **10%**.

.question[
- _Head of Serv. of Surgery at Hospital da Guarda:_<br> Thanks to the practices we implemented, we only had 30 complicated cases in the year 2022.
]
- His thesis: this proves that the improvements he implemented led to a reduction in the probability of serious complications occurring at the Hospital da Guarda, compared to the rest of the country.
--
**Could it be?**

---

### Data: Exercise

- The total number of surgeries at Hospital da Guarda in 2022 was 500.
- Create from scratch, using code, a table with a record for each surgery and a binary variable that indicates if there were complications. Tips:
  - First, create a table with a variable indicating there were no complications in all records
  - Second, change the value of only the first 30 lines
--

```{r}
surgeries_Guarda <- tibble(
   no_surgery = seq(0, 500),
   I_Complications = FALSE
)
# Binary variable => use logical type values
surgeries_Guarda$I_Complications[0:30] = TRUE

```

---

### Data: Exercise 2

- How could you perform the same procedure above (only changing those values to TRUE), but using a tidyverse function?

--

```{r}
surgeries_Guarda <- surgeries_Guarda %>%
   mutate(result = if_else(no_surgery <= 30, TRUE, FALSE))

# Bonus: without using the surgery number variable
surgeries_Guarda %>% mutate(result = if_else(n() <= 30, TRUE, FALSE))

```



---

## Parameter and measurement or estimate


$p~$: probability of serious complications occurring in surgeries in Portugal $= 10\%$.



$\hat{p}~$: percentage of complications in the Guard in 2022 (in the sample) = $\frac{30}{500} = 6\%$

--

.question[
Is it possible to confirm the director's statement with these data?
]

--

- No -- **correlation vs. causality**: the probability of complications in Guarda may be different from the rest of the country for reasons other than the changes promoted
   - Even so, we can assess if the rate obtained, 6%, really means that the probability in Guarda is lower  compared to the rest of the country.
   - Or if on the contrary, in 2022 the rate was lower by mere chance.
---

## 2 hypotheses

Hypothesis testing begins by defining the null and alternative hypotheses:

- **Null hypothesis:** "There is no difference between Guarda and the rest of the country"

--

- **Alternative hypothesis:** "The rate is actually lower in Guarda"

---

## A hypothesis test is like a trial in court

- **Null hypothesis**, $H_0$: The defendant is innocent

- **Alternative hypothesis**, $H_A$: Defendant is guilty

--

- **Present the evidence:** Collect the *data*

--

**Judge based on evidence:**

   - If the null hypothesis is true:
     - (That is, if the probability of complications remains the same as in the rest of the country)
     - Is it plausible that such a low value was obtained this year?

.center[ **Yes**: H0 vs. **No**: H0 is rejected ]
    
---

## Structure of any hypothesis test

- You start with the null hypothesis, $H_0$, which represents the status quo
- The alternative hypothesis, $H_A$, represents the research question, i.e. the new information, whose likelihood we want to test.
   - To test the hypothesis that $H_0$ is true, obtain the **p-value** $\rightarrow$ probability of a result as  *extreme* than the one observed or more, under the null hypothesis
- Two possibilities:

.pull-left[**high p-value** $\Rightarrow$ proofs <u>don't</u> allow rejecting $H_0$
- The observed complication rate (6%) seems plausible under $H_0$]
.pull-right[
**low p-value** $\Rightarrow$ proofs allow rejecting $H_0$
- Under $H_0$, it seems difficult to observe such a low rate
]

